The XAI Paradox: Systems that Perform Well for the Wrong Reasons

### 2018-08-28 Paper.tex
Div

\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (The XAI Paradox: Systems that Perform Well for the Wrong Reasons)
/Author (Cor Steging, Lambert Schomaker, Bart Verheij)}
\setcounter{secnumdepth}{0}  
 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{The XAI Paradox: Systems that Perform Well for the Wrong Reasons}
\author{Cor Steging, Lambert Schomaker, Bart Verheij\\
Department of Artificial Intelligence, Bernoulli Institute\\
University of Groningen\\
}
\maketitle
\begin{abstract}
Many of the successful modern machine learning approaches can be described as ``black box'' systems; these systems perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that are able to explain to their users why they made a particular decision. Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. Furthermore, when multiple combined conditions define a dataset, systems can perform well on the combined problem and not learn each of the individual conditions. Instead, a confounding structure within the data is learnt that still allows making the correct decisions. With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
%Many of the successful modern machine learning approaches can be described as ``black box'' systems; these systems perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that are able to explain to their users why they made a particular decision. Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of neural networks and decisions trees that perform well are not necessarily sound. Furthermore, when multiple conditions define a dataset, systems can perform well on the combined problem yet not learn each of the individual conditions that make up the combined problem. Instead, a confounding structure within the data is learnt that still allows making the correct decisions. With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
\end{abstract}

\section{Introduction}
In recent years, the notion of artificial intelligence has reached the masses with the introduction of smart devices, digital assistants, self-driving cars and many other inventions. This booming interest in AI can largely be attributed to the success of machine learning in combination with the growing availability of big data. Despite their success, many of the most commonly employed machine learning techniques inherently lack a possibility to provide an explanation of the way in which they reason. These ``black box'' systems are often distrusted by their users because they do not explain why particular decisions are made (cf. also the discussion about a `right to an explanation' in the EU General Data Protection Act; see e.g. \cite{edwards2017slave}).

The response to the lack of transparency is the emerging field of Explainable Artificial Intelligence (XAI) \cite{gunning2017explainable}. XAI aims to create explainable models that are able to explicitly describe the inner workings of a ``black box'' machine learning system, and thus provide the essential explanations to their users. Examples of such XAI systems are rule extraction algorithms, such as \cite{lu2017neurorule}, that create  sets of rules for the users that describe the reasoning of ``black box'' systems, or glass-box systems, like \cite{holzinger2017glass}, in which the users are able to influence what the systems learn. Yet another approach is used in a recent deep learning system for diagnosing retinal diseases (outperforming medical experts). The system is divided into a framework of smaller systems; one for each stage of the diagnostic process \cite{de2018clinically}, making it easier for the clinicians to investigate and understand the reasoning of the system. In these and different ways, XAI systems aim to yield accurate and meaningful explanations of their reasoning, without sacrificing performance.

A question that must be asked in XAI research is whether or not the explanation that a XAI system gives indeed makes sense in the domain. That this is not a trivial issue is in particular suggested by the phenomenon of adversarial examples in image classification, where input images are slightly altered using a perturbation \cite{yuan2017adversarial}. To the human eye, there is no apparent difference between the original input image and its altered version, while a machine learning system makes a completely different classification after the image is altered. This suggests that the understanding that the systems have differs strongly from that of humans, despite the fact that the system usually makes the correct classifications. In other words, systems perform well, but apparently for the wrong reasons. In and of itself, this is still impressive, as the machine learning system is able to extrapolate structures from the data that we as humans are not able to perceive, and consequently achieve a high performance. As a result, however, we can expect that when an explainable model is generated from these systems using XAI techniques, explanations that are generated may not make sense to the users, countering the desired transparancy and decreasing trust.

The aim of this study is to investigate this question: to what extent do systems learn structures that make sense in a domain? In order to operationalize this idea, we generate structured datasets, where the embedded structure is a model of what makes sense in a domain. We investigate machine learning techniques in terms of how well they are able to learn the structure that defines datasets, rather than only in terms of a general performance accuracy. More specifically, artificial datasets will be generated from a set of predetermined formal conditions in terms of symmetric Boolean functions, on which neural networks and decision tree systems will train. In this way, we can study how well the systems are able to extrapolate the correct conditions from the data. By combining multiple conditions defining datasets, we can investigate interaction effects whether the system learns each of the individual conditions correctly, or instead learns a different confounding structure that still accurately maps the input to the output.

Our method of using artificial datasets with predetermined structure was inspired by neural network research in a legal setting \cite{bench1993neural}, a 1990s study of XAI themes in the field of AI \& Law. In that field, XAI issues arose already then, since explainability and transparency are especially pertinent in the law, where decisions must be made for the right reasons. That work used a specific legal example. We present an abstract, parameterized set of problems (defined using families of symmetric Boolean functions) in order to gain more insights about the task of learning structure that makes sense in a domain. Before describing our specific method and results, we discuss the findings of \cite{bench1993neural}.

\section{Previous Research:  \\the Welfare Benefit Dataset}
\cite{bench1993neural} uses a particular case study in the field of law investigating whether neural networks can achieve a high performance on open texture law problems. In that work, it is examined whether or not the rationale that the neural networks use in their classification is acceptable. To this end, a fictional dataset is generated that describes the personal information of elderly people and whether they are eligible for a particular welfare benefit. The eligibility depends on the six conditions shown below, and the personal information of the individual. If and only if all of the six conditions apply to an individual, her or she is eligible for a benefit.

@@

\noindent Neural networks are trained on this dataset and tasked with predicting the eligibility of new individuals. High classification accuracies are achieved by these neural networks, with the number of correct predictions ranging from 98.75\% to 99.25\%.

@@

\section{Method: Symmetric Boolean Functions}
We saw that in the welfare benefit dataset the systems were unable to learn all of the conditions that defined the data, despite achieving a high classification accuracy. Individually, the conditions were quite simple and all modern machine learning systems should be able learn them. Combined, however, they become more difficult to learn. This could indicate that there might be a type of interaction effect between conditions, that causes a condition to be learned less successfully when other conditions are present. For instance, a machine learning algorithm may be able to easily learn two conditions separately, but have difficulty when attempting to learn both at the same time. This study sets out to investigate how well machine learning systems are able to internalize such conditions that define a training dataset. Furthermore, the possible interaction effect between conditions will be examined.

We use %The type of functions that neural networks and other machine learning algorithms have been shown to struggle with in previous research can be generalized into 
symmetrical Boolean functions. The output of these functions is only based on the number of 1's in the input vector, hence the location of the 1's in the input vector is irrelevant. An example of a symmetrical Boolean function with two variables is the XOR function (well-known for its role in the history of neural networks), which provides an output of 1 if and only if the number of 1's in the input vector is equal to 1. With more than two variables, the XOR function can be generalized to the parity function, which is a function that returns an output of 1 if and only if there is an odd number of 1's in the input vector. We use three families of symmetrical Boolean functions \cite{wegener1987complexity}:

@@

\subsection{Datasets}
%%%BV VERHELDEREN
For our experiment, a parametrized series of datasets is generated based on conditions defined in terms of symmetrical Boolean functions. Each instance in the dataset has an output value, which is true if and only if all of its conditions are true as well. This set of conditions can include the M-out-of-N function, the exact value function and the parity function. Each condition also has a set of parameters. As described earlier, parameter $n$ denotes the number of variables of the condition and parameter $m$ is used in M-out-of-N functions and exact value functions to determine the value of the output. The total number of variables of each instances in the datasets is therefore based on the number of variables of each of the conditions. Table \ref{tbl:instance_example} shows two example instances of such a dataset. The instance of this dataset consist of two conditions: the parity condition with 2 variables (in red) and the M-out-of-N condition with 3 variables (in blue), where parameter $m$ is set to 2. In the first instance, the parity condition is false and the M-out-of-N condition is true. The output is therefore false, because all conditions  of an instance need to be true in order for the output to be true. In the second instance, the output is true, as both the parity condition (red) and the M-out-of-N condition (blue) are true.

\begin{table}[t]
\caption{Two example instances of a dataset with 2 parity variables (red) and 3 M-out-of-N variables (blue), where $m$=2.}
\centering
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\cellcolor[HTML]{FD6864}A & \cellcolor[HTML]{FD6864}B & \cellcolor[HTML]{34CDF9}C & \cellcolor[HTML]{34CDF9}D & \cellcolor[HTML]{34CDF9}E & Output \\ \hline
1                         & 1                         & 0                         & 1                         & 1                         & 0      \\ \hline
\end{tabular}
\centering
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\cellcolor[HTML]{FD6864}A & \cellcolor[HTML]{FD6864}B & \cellcolor[HTML]{34CDF9}C & \cellcolor[HTML]{34CDF9}D & \cellcolor[HTML]{34CDF9}E & Output \\ \hline
1                         & 0                         & 0                         & 1                         & 1                         & 1      \\ \hline
\end{tabular}
\label{tbl:instance_example}
\end{table}

For any set of conditions, the following datasets are generated: a training set, a general test set and a specific test set for each condition. First of all, a training set consists of 150,000 instances where half of the instances are generated randomly such that their output value is true, and the other half are generated randomly such that their output value is false.  In the latter half, the instances are generated such that they fail on each of its conditions equally. With two conditions for example, this means that the latter half (the half where the output is false) would consists for 50\% out of instances where the first condition is false and for 50\% out of instances where the second condition is false. If a condition is not specifically set up to fail, its variables are provided with random values. The general test set consists of 150,000 instances, and is generated in the exact same fashion.

For each condition, an additional specific test set is generated. This test set is used to determine how well a system is able learn a specific condition. Given such a condition, its test set is constructed by generating all of the possible input values that the condition can have. For example, the XOR condition (or parity function with $n$=2) has $2^2=4$ possible values and would therefore have a specific test set with 4 instances. The variables of any additional conditions are given random values such that these conditions are satisfied. This ensures that the output is dependent on only the single condition for which the test set is generated. This allows us to measure how well that particular condition is learned by the system.

@@

%%%BV INCLUDE A SUBSECTION `Experiments' HERE where you describe all experiments, including noise
\subsection{Experiments}
Both machine learning systems are first trained and tested using datasets with only a single condition; either the parity, M-out-of-N or exact value function. This will determine whether the systems are actually able to learn each of the conditions individually. The systems will then be trained and tested on datasets with two conditions, such as the example instances in Table \ref{tbl:instance_example}, in order to investigate both how well the systems are able to learn the combined problem and how well they are able to learn the individual conditions that make up the combined problem.

To show the ability of the systems perform under noise, a additional tests will be performed. In these experiments, a particular percentage of the instances of the training datasets are given a ``noisy variable''. These noisy variables have their values inverted, such that a 1 becomes a 0 and a 0 becomes a 1. After training on the datasets with noise, the systems are tasked with classifying the same test sets as before.

@@

\subsection{Single Condition Datasets}
First of all, datasets are generated with only one condition with different numbers of variables $n$; from 2 to 18.  For the M-out-of-N and exact value condition, the value of $m$ will be varied as well: between 1 and $n$. For each condition and each possible combination of $m$ and $n$, both the decision tree algorithm and the neural network are trained on the training dataset and then tested on the general test set and the specific condition test set. In Figure \ref{fig:DTvsNN} the accuracy of the decision tree and the neural network on the general test set is shown versus the number of variables for all three conditions. For the M-out-of-N and exact value condition, the mean accuracies across all possible values values of $m$ are displayed.

@@

Varying the percentage of instances with a noisy variable between 0 and 100\% yields the graphs shown in Figure \ref{fig:SBF_noise}. These graphs display the mean accuracies of the decision tree (red) and the neural network (blue) on the specific test set for each of the three symmetrical Boolean functions.  The accuracies are average accuracies across values of $n$ ranging from 2 to 10 and across all possible values of $m$. It should be noted that inverting the value of a variable does not always mean that the outcome of the instance becomes incorrect. Therefore if 50\% of the instances have a noisy variable, it does not necessarily mean that 50\% of the instances have an incorrect output.

@@

Even under extreme levels of noise, the systems are still able to learn the M-out-of-N and exact value function quite effectively, showing the robustness of machine learning results.

%%%CS Leave this out entirely, or move it to discussion?
\iffalse
The M-out-of-N function is most robust to noise, as seen in Figure \ref{fig:SBF_MU}. This is because inverting one variable of the input vector only changes the output in two cases. First of all, if the total number of 1's is equal to $m$ and the output is true, changing a 1 to a 0 due to noise should change the output to false. Secondly, if the total number of 1's is equal to $m-1$ and the output is false, and the noise changes a variable from 0 to 1, the output should be true.

The accuracies on the exact function drop faster with more noise, as seen in Figure \ref{fig:SBF_EU}. When an instance of this dataset has a true output, and one of its variables is inverted, the total number of 1's will no longer be equal to $m$, and therefore the output will be incorrect. If an instance in the exact dataset has an output that is false, however, the noise can do two things: the number of 1's might be equal to $m$, thus creating a true output, or (the more likely alternative) the number of 1's may still not be equal to $m$, and the output is still false. This effect can be seen in Figure \ref{fig:SBF_EU}, as the systems are only able to correctly classify the incorrect instances, thus achieving an accuracy of roughly 50\%.

The parity function is a different case, as its output is dependent on whether the amount of 1's in the input vector is odd or even. By adding a noisy variable to an instance of the parity function, the amount of 1's will always change from odd to even, or even to odd. This means that every instance of a pairty dataset that has a noisy variable automatically has the wrong output. This is why the accuracies on the parity function drops more significantly than on the other functions (to 0\% in the case of the decision tree in Figure \ref{fig:SBF_PU}). 
\fi

@@

\begin{table}[h!]
\centering
\caption{The mean accuracies of the decision tree on each test set after training on each interaction.}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{|c|c|c|c|}
\hline 
 & General &  Function 1 & Function 2 \\ 
 & accuracy &  accuracy & accuracy \\ 
\hline 
Parity, Exact &  98.50 & 17.48 & 78.75 \\ 
\hline 
Parity, M-out-of-N  & 99.99 & 30.36 & 81.75 \\ 
\hline 
M-out-of-N, Exact & 99.99 & 96.00 & 32.53 \\ 
\hline 
\end{tabular} %
}
\label{tbl:interactions_DT}
\end{table}

\begin{table}[h!]
\centering
\caption{The mean accuracies of the neural network on each test set after training on each interaction. }
\resizebox{\columnwidth}{!}{%
\begin{tabular}{|c|c|c|c|}
\hline 
 & General &  Function 1 & Function 2 \\ 
 & accuracy &  accuracy & accuracy \\ 
\hline 
Parity, Exact & 99.65 & 14.59 & 78.31 \\ 
\hline 
Parity, M-out-of-N & 100.00 & 30.59 & 81.65 \\ 
\hline 
M-out-of-N, Exact & 99.99 & 99.88 & 26.14  \\ 
\hline 
\end{tabular}%
} 
\label{tbl:interactions_NN}
\end{table}

An overview of the average accuracies of the decision trees and neural networks is given in Table \ref{tbl:interactions_DT} and Table \ref{tbl:interactions_NN} respectively. These show the mean accuracies for each interaction on each of the different test sets: the general test set, and one test specific test set for each of the two functions of the interaction. What is clear from these tables, is that the mean accuracy of both the neural network and the decision tree on the general test set is quite high for each interaction. However, the accuracies on the specific test sets, which indicate how well the functions themselves are learned, are almost always a lot lower. With the parity-exact interaction, for instance, the decision tree algorithm preforms with an accuracy of 98.5\% on the general test set, but only 17.5\% on the parity test set and 78.8\% on the exact test set. This is shows that a system can perform with a high accuracy, even if it has not learned all of the conditions correctly. The accuracies of the neural network and the decision tree system are also quite similar to each other, which shows that the trends in accuracies are not limited to a single machine learning algorithm. The relationships between the number of variables for each function and the accuracies of each of the tests sets are recorded as well, an example of which can be seen in the heatmaps of Figure \ref{fig:Heatmap_DT_ME} and Figure \ref{fig:Heatmap_NN_ME}. These display the mean accuracies of the decision tree on the three different test sets for the M-out-of-n and exact value function interaction versus the number of variables of both functions on the x- and y-axis.

@@

We have investigated these effects for all combinations of the conditions. A summarized overview of the effects that the conditions have in the other interactions is shown in Table \ref{tbl:highlow}.

\begin{table}[h!]
\centering
\caption{This table shows the  effect that a high number of variables of a function has on the performance of a system (in terms of learning the function) for all functions that it interacts with.}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{|c|c|c|c|}
\hline 
Performance on $\downarrow$ & Parity & Exact & M-out-of-N \\ 
\hline 
Parity & - & decreases & decreases \\ 
\hline 
Exact & increases & - & decreases \\ 
\hline 
M-out-of-N & increases & decreases & - \\ 
\hline 
\end{tabular}%
}
\label{tbl:highlow}
\end{table}

>>>>

\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (The XAI Paradox: Systems that Perform Well for the Wrong Reasons)
/Author (Cor Steging, Lambert Schomaker, Bart Verheij)}
\setcounter{secnumdepth}{0}  
 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{The XAI Paradox: Systems that Perform Well for the Wrong Reasons}
\author{Cor Steging, Lambert Schomaker, Bart Verheij\\
Department of Artificial Intelligence, Bernoulli Institute\\
University of Groningen\\
}
\maketitle
\begin{abstract}
Successful machine learning approaches are often regarded as ``black box'' systems: they perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that can provide explanations underlying their internal reasoning. 
In this study, we investigate to
what extent systems learn explanatory structures that make sense in
a domain. 
Using artificial datasets whose internal structure is determined beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. 
We generate datasets from predetermined formal conditions defined in terms of symmetric Boolean functions. When using datasets generated by conjunctions of conditions, we find that trained systems perform well on the conjunction problem, but---paradoxically---not on the individual conditions separately. This implies that a confounding
%Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. In particular, when multiple combined conditions define a dataset, systems can perform well on the combined problem and not learn each of the individual conditions. Instead, a confounding 
structure within the data is learnt that still allows making the correct decisions. In other words, the systems perform well, but for the wrong reasons, providing a puzzle for the XAI movement.
%With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
%Many of the successful modern machine learning approaches can be described as ``black box'' systems; these systems perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that are able to explain to their users why they made a particular decision. Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of neural networks and decisions trees that perform well are not necessarily sound. Furthermore, when multiple conditions define a dataset, systems can perform well on the combined problem yet not learn each of the individual conditions that make up the combined problem. Instead, a confounding structure within the data is learnt that still allows making the correct decisions. With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
\end{abstract}

\section{Introduction}
In recent years, the notion of artificial intelligence has reached the masses with the introduction of smart devices, digital assistants, self-driving cars and many other inventions. This booming interest in AI can largely be attributed to the success of machine learning in combination with the growing availability of big data. Commonly employed machine learning techniques are inherently ``black box'' systems, lacking a possibility to provide an explanation underlying their reasononing.

In certain domains, such as the law and medicine, there is a need for more explainability and transparency (cf. also the discussion about a `right to an explanation' in the EU General Data Protection Act; see e.g. \cite{edwards2017slave}).
Increasing explainability and transparency is studied in the emerging field of Explainable Artificial Intelligence (XAI) \cite{gunning2017explainable}. 
%XAI aims to create explainable models that are able to explicitly describe the inner workings of a ``black box'' machine learning system, and thus provide the essential explanations to their users. 
Examples of such XAI systems are rule extraction algorithms, such as \cite{lu2017neurorule}, that create  sets of rules for the users that describe the reasoning of ``black box'' systems, or glass-box systems, like \cite{holzinger2017glass}, in which the users are able to influence what the systems learn. Yet another approach is used in a recent deep learning system for diagnosing retinal diseases (outperforming medical experts). The system is divided into a framework of smaller systems; one for each stage of the diagnostic process \cite{de2018clinically}, making it easier for the clinicians to investigate and understand the modular reasoning of the system. In these and different ways, XAI systems aim to yield accurate and meaningful explanations of their reasoning, without sacrificing performance.

A question that must be asked in XAI research is whether or not the explanation that a XAI system gives indeed makes sense in the domain. That this is not a trivial issue is in particular suggested by the phenomenon of adversarial examples in image classification, where input images are slightly altered using a perturbation \cite{yuan2017adversarial}. To the human eye, there is no apparent difference between the original input image and its altered version, while a machine learning system makes a completely different classification after the image is altered. This suggests that the understanding that the systems have differs strongly from that of humans, despite the fact that the system usually makes the correct classifications. In other words, systems perform well, but apparently for the wrong reasons. In and of itself, this is still impressive, as the machine learning system is able to extrapolate structures from the data that we as humans are not able to perceive, and consequently achieve a high performance. However, explanations that are generated using XAI techniques may not make sense to the users, countering the goals of transparancy and explainability.

The aim of this study is to investigate this question: to what extent do systems learn structures that make sense in a domain? In order to operationalize this idea, we generate structured datasets, where the embedded structure is used as a model of what makes sense in a domain. We investigate machine learning techniques in terms of how well they are able to learn the structure that defines datasets, rather than only in terms of a general performance accuracy. More specifically, artificial datasets will be generated from a set of predetermined formal conditions in terms of symmetric Boolean functions, on which neural networks and decision tree systems will train. In this way, we can study how well the systems are able to learn the correct conditions from the data. 
By using conjunctions of conditions when defining datasets, 
we can investigate whether the system learns each of the individual conditions correctly, or instead learns a different confounding structure that still accurately maps the input to the output.

Our method of using artificial datasets with predetermined structure was inspired by neural network research in a legal setting \cite{bench1993neural}, a 1990s study of XAI themes in the field of AI \& Law. In that field, XAI issues arose already then, since explainability and transparency are especially pertinent in the law, where decisions must be made for the right reasons. That work used a specific legal example. In this study, we use an abstract, parameterized set of problems (defined using families of symmetric Boolean functions) in order to investigate the task of learning structure that makes sense in a domain. %Before describing our specific method and results, we discuss the findings of \cite{bench1993neural}.

\section{Previous Research:  \\the Welfare Benefit Dataset}
\cite{bench1993neural} uses a particular case study in the field of law investigating whether neural networks can achieve a high performance on open texture law problems. He examines whether or not the rationale that the neural networks use in their classification is acceptable. To this end, a fictional dataset is generated that describes the personal information of elderly people and whether they are eligible for a particular welfare benefit. The eligibility depends on the six conditions shown below, and the personal information of the individual. If and only if all of the six conditions apply to an individual, her or she is eligible for a benefit.

@@

\noindent Training neural networks on this dataset for predicting the eligibility of new individuals leads to high classification accuracies, with the number of correct predictions ranging from 98.75\% to 99.25\%.

@@

\section{Method: Symmetric Boolean Functions}
We saw that in the welfare benefit dataset systems were unable to learn all of the conditions defining the data, despite achieving a high classification accuracy. Individually, the conditions were quite simple and all modern machine learning systems should be able learn them. Combined, however, they become more difficult to learn. This could indicate that there is an interaction effect between conditions, that causes a condition to be learned less successfully when other conditions are present. 
%For instance, a machine learning algorithm may be able to easily learn two conditions separately, but have difficulty when attempting to learn both at the same time. 
This study sets out to investigate how well machine learning systems are able to internalize such conditions that define a training dataset. Furthermore, the possible interaction effect between conditions will be examined.

For defining conditions that generate datasets, we use %The type of functions that neural networks and other machine learning algorithms have been shown to struggle with in previous research can be generalized into 
symmetrical Boolean functions. The output of these functions is only based on the number of 1's in the input vector, hence the location of the 1's in the input vector is irrelevant. An example of a symmetrical Boolean function with two variables is the XOR function (well-known for its role in the history of neural networks), which provides an output of 1 if and only if the number of 1's in the input vector is equal to 1. With more than two variables, the XOR function can be generalized to the parity function, which is a function that returns an output of 1 if and only if there is an odd number of 1's in the input vector. We use three families of symmetrical Boolean functions \cite{wegener1987complexity}:

@@

\subsection{Datasets}
%%%BV VERHELDEREN
For our experiments, a parametrized series of datasets is generated based on conditions defined in terms of symmetrical Boolean functions. Each instance in the dataset has an output value, which is true if and only if all conditions are true as well. This set of conditions can include the M-out-of-N function, the exact value function and the parity function, with parameters $n$ and $m$ as above. Parameter $n$ denotes the number of variables of the condition and parameter $m$ is used in M-out-of-N functions and exact value functions to determine the value of the output. 
The total number of variables of each instance in the datasets is based on the number of variables of each of the conditions, as for instance in the examples in Table \ref{tbl:instance_example}. The output variable is set to true if and only if two conditions both hold: the parity condition with 2 variables A and B (in red) and the M-out-of-N condition with 3 variables C, D and E (in blue), where parameter $m$ is set to 2. In the first instance, the parity condition is false and the M-out-of-N condition is true; hence output is false. In the second instance, the output variable is true, as both the parity condition and the M-out-of-N condition are true.

\begin{table}[b]
\caption{Two example instances of a dataset with 2 parity variables A and B (red) and 3 M-out-of-N variables C, D and E (blue), where $m$=2.}
\centering
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\cellcolor[HTML]{FD6864}A & \cellcolor[HTML]{FD6864}B & \cellcolor[HTML]{34CDF9}C & \cellcolor[HTML]{34CDF9}D & \cellcolor[HTML]{34CDF9}E & Output \\ \hline
1                         & 1                         & 0                         & 1                         & 1                         & 0      \\ \hline
%\end{tabular}
%\centering
%\begin{tabular}{|l|l|l|l|l|l|}
%\hline
%\cellcolor[HTML]{FD6864}A & \cellcolor[HTML]{FD6864}B & \cellcolor[HTML]{34CDF9}C & \cellcolor[HTML]{34CDF9}D & \cellcolor[HTML]{34CDF9}E & Output \\ \hline
1                         & 0                         & 0                         & 1                         & 1                         & 1      \\ \hline
\end{tabular}
\label{tbl:instance_example}
\end{table}

For any pair of conditions (Parity, Exact; Parity, M-out-of-N; M-out-of-N, Exact), the following datasets are generated: a training set, a general test set and a specific test set for each condition. First of all, a training set consists of 150,000 instances where half of the instances are generated randomly such that their output value is true, and the other half are generated randomly such that their output value is false.  In the latter half, the instances are generated such that they fail on each of its conditions equally. With two conditions for example, this means that the latter half (the half where the output is false) would consists for 50\% out of instances where the first condition is false and for 50\% out of instances where the second condition is false. If a condition is not specifically set up to fail, its variables are provided with random values. The general test set consists of 150,000 instances, and is generated in the exact same fashion.

For each condition, also a specific test set is generated, used to determine how well a system is able to learn that specific condition. Given such a condition, its test set is constructed by generating all of the possible input values that the condition can have. For example, the XOR condition (or parity function with $n$=2) has $2^2=4$ possible values and would therefore have a specific test set with 4 instances. The variables of any additional conditions are given random values such that these conditions are satisfied. This ensures that the output is dependent on only the single condition for which the test set is generated. %This allows us to measure how well that particular condition is learned by the system.

@@

%%%BV INCLUDE A SUBSECTION `Experiments' HERE where you describe all experiments, including noise
\subsection{Experiments}
Both machine learning algorithms are first trained and tested using datasets with only a single condition; either the parity, M-out-of-N or exact value function. This will determine whether the systems are actually able to learn each of the conditions individually. The systems will then be trained and tested on datasets for each pair of conditions, in order to investigate both how well the systems are able to learn the combined problem and how well they are able to learn the individual conditions that make up the combined problem.

To show the ability of the systems perform under noise, additional tests will be performed. In these experiments, a particular percentage of the instances of the training datasets are given a ``noisy variable''. These noisy variables have their values inverted, such that a 1 becomes a 0 and a 0 becomes a 1. After training on the datasets with noise, the systems are tasked with classifying the same test sets as before.

@@

\subsection{Single Condition Datasets}
First of all, datasets are generated with only one condition with different numbers of variables $n$; from 2 to 18.  For the M-out-of-N and exact value condition, the value of $m$ varies as well: between 1 and $n$. For each condition and each possible combination of $m$ and $n$, both the decision tree algorithm and the neural network are trained on the training dataset and then tested on the general test set and the specific condition test set. In Figure \ref{fig:DTvsNN} the accuracy of the decision tree and the neural network on the general test set is shown versus the number of variables for all three conditions. For the M-out-of-N and exact value condition, the mean accuracies across all possible values values of $m$ are displayed.

@@

Varying the percentage of instances with a noisy variable between 0 and 100\% yields the graphs shown in Figure \ref{fig:SBF_noise}. These graphs display the mean accuracies of the decision tree (red) and the neural network (blue) on the specific test set for each of the three symmetric Boolean functions.  The accuracies are average accuracies across values of $n$ ranging from 2 to 10 and across all possible values of $m$. It should be noted that inverting the value of a variable does not always make the outcome of the instance incorrect, and 50\% instances with a noisy variable need not mean 50\% incorrect instances.

@@

The results show that even under extreme levels of noise, the systems are able to learn the M-out-of-N and exact value function quite effectively.
%, showing the robustness of machine learning results. 
%%%CS Leave this out entirely, or move it to discussion?
\iftrue
The M-out-of-N function is most robust to noise, as seen in Figure \ref{fig:SBF_MU}. %This is because inverting one variable of the input vector only changes the output in two cases. First of all, if the total number of 1's is equal to $m$ and the output is true, changing a 1 to a 0 due to noise should change the output to false. Secondly, if the total number of 1's is equal to $m-1$ and the output is false, and the noise changes a variable from 0 to 1, the output should be true.  
The accuracies on the exact function drop faster with more noise, as seen in Figure~\ref{fig:SBF_EU}. 
%When an instance of this dataset has a true output, and one of its variables is inverted, the total number of 1's will no longer be equal to $m$, and therefore the output will be incorrect. If an instance in the exact dataset has an output that is false, however, the noise can do two things: the number of 1's might be equal to $m$, thus creating a true output, or (the more likely alternative) the number of 1's may still not be equal to $m$, and the output is still false. This effect can be seen in Figure \ref{fig:SBF_EU}, as the systems are only able to correctly classify the incorrect instances, thus achieving an accuracy of roughly 50\%.

%The parity function is a different case, as its output is dependent on whether the amount of 1's in the input vector is odd or even. By adding a noisy variable to an instance of the parity function, the amount of 1's will always change from odd to even, or even to odd. This means that every instance of a pairty dataset that has a noisy variable automatically has the wrong output. This is why the accuracies on the parity function drops more significantly than on the other functions (to 0\% in the case of the decision tree in Figure \ref{fig:SBF_PU}). 
\fi

@@

\begin{table}[b!]
\centering
\caption{The mean accuracies of the decision tree on each test set after training on each interaction.}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{|c|c|c|c|}
\hline 
 & General &  Function 1 & Function 2 \\ 
 & accuracy &  accuracy & accuracy \\ 
\hline 
Parity, Exact &  98.50 & 17.48 & 78.75 \\ 
\hline 
Parity, M-out-of-N  & 99.99 & 30.36 & 81.75 \\ 
\hline 
M-out-of-N, Exact & 99.99 & 96.00 & 32.53 \\ 
\hline 
\end{tabular} %
}
\label{tbl:interactions_DT}
\end{table}

\begin{table}[b!]
\centering
\caption{The mean accuracies of the neural network on each test set after training on each interaction. }
\resizebox{\columnwidth}{!}{%
\begin{tabular}{|c|c|c|c|}
\hline 
 & General &  Function 1 & Function 2 \\ 
 & accuracy &  accuracy & accuracy \\ 
\hline 
Parity, Exact & 99.65 & 14.59 & 78.31 \\ 
\hline 
Parity, M-out-of-N & 100.00 & 30.59 & 81.65 \\ 
\hline 
M-out-of-N, Exact & 99.99 & 99.88 & 26.14  \\ 
\hline 
\end{tabular}%
} 
\label{tbl:interactions_NN}
\end{table}

An overview of the average accuracies of the decision trees and neural networks is given in Tables \ref{tbl:interactions_DT} and \ref{tbl:interactions_NN}. These show the mean accuracies for each interaction on each of the different test sets: the general test set, and one test specific test set for each of the two functions of the interaction. What is clear from these tables, is that the mean accuracy of both the neural network and the decision tree on the general test set is quite high for each interaction. However, the accuracies on the specific test sets, which indicate how well the functions themselves are learned, are almost always a lot lower. With the parity-exact interaction, for instance, the decision tree algorithm preforms with an accuracy of 98.5\% on the general test set, but only 17.5\% on the parity test set and 78.8\% on the exact test set. This shows that a system can perform with a high accuracy, even if it has not learned all of the conditions correctly. The accuracies of the neural network and the decision tree system are also quite similar to each other, which shows that the trends in accuracies are not limited to a single machine learning algorithm. The relationships between the number of variables for each function and the accuracies of each of the tests sets are recorded as well, an example of which can be seen in the heatmaps of Figure \ref{fig:Heatmap_DT_ME} and Figure \ref{fig:Heatmap_NN_ME}. These display the mean accuracies of the decision tree on the three different test sets for the M-out-of-n and exact value function interaction versus the number of variables of both functions on the x- and y-axis.

@@

We have investigated these effects for all pairs of conditions. A summarized overview of the effects that the conditions have in the other interactions is shown in Table \ref{tbl:highlow}.

\begin{table}[b!]
\centering
\caption{This table shows the  effect that a high number of variables of a function has on the performance of a system (in terms of learning the function) for all functions that it interacts with.}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{|c|c|c|c|}
\hline 
Performance on $\downarrow$ & Parity & Exact & M-out-of-N \\ 
\hline 
Parity & - & decreases & decreases \\ 
\hline 
Exact & increases & - & decreases \\ 
\hline 
M-out-of-N & increases & decreases & - \\ 
\hline 
\end{tabular}%
}
\label{tbl:highlow}
\end{table}
